Human capital mediates natural selection in contemporary humans

### 2022-06-09 why-natural-selection.Rmd
Changes for final revision:   - use fewer controls in mediation analysis   - only show confidence intervals in mediation analysis plot when they are correct   - tiny colour tweaks

```{r plot-res-wt, fig.cap = "Selection effects: weighted and unweighted regressions.  Each point represents a single bivariate regression of number of children on a polygenic score. P value threshold is 0.05, Bonferroni-corrected for multiple comparisons. Confidence intervals are uncorrected."}

@@

res_mediation %<>% mutate(
                     # 0.025 not 0.05 because two-sided
                     sig     = abs(statistic_ind) > qnorm(1 - 0.025/
                                                            nrow(res_mediation)),
                     pos_sig = sig & prop_ind > 0
                   )

@@

```{r plot-mediation, fig.align = "center", fig.cap = "Proportion of selection effect mediated by educational attainment, among polygenic scores with significant selection effects"}

res_mediation %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, prop_ind, order_abs(1)),
        `Indirect effect p < 0.05/23` = sig 
      ) %>% 
      ggplot(aes(y = term)) + 
        geom_col(aes(x = prop_ind, fill = `Indirect effect p < 0.05/23`), 
                   width = .5) + 
        geom_errorbar(aes(
                          xmin = prop_ind_conf_low, 
                          xmax = prop_ind_conf_high,
                          linetype = "Bootstrap 95% c.i. uncorrected",
                        ),
                        width = 0.2,
                        color = "grey50"
                      ) +
        scale_fill_manual(values = c("FALSE" = "grey65", "TRUE" = "darkgreen")) +
        my_vline +
        scale_x_continuous(labels = scales::percent, n.breaks = 7) +
        coord_cartesian(xlim = c(-0.5, 1)) +
        labs(x = "Proportion mediated", y = "") +
        scale_linetype_manual(name = "", 
          values = c("Bootstrap 95% c.i. uncorrected" = 1)) +
        my_theme(
          panel.grid.major.y = element_blank(),
          panel.grid.minor.x = element_blank(),
          legend.justification = "top"
        )

@@

Because selection effects are concentrated in lower-income groups, they may also
increase inequality with respect to polygenic scores. For example, Figure
\@ref(fig:plot-mean-EA3-by-income) plots mean polygenic scores for educational
attainment (EA3) among children from households of different income groups. The
green bars show the actual means, i.e. parents' mean polygenic score weighted by
number of children. The grey bars show the hypothetical means if all households
had equal numbers of children. Natural selection against genes associated with
educational attainment is stronger at the bottom of the income distribution, and
this increases the differences between groups. Overall, natural selection
increases the correlation of polygenic scores with income for `r n_increases`
out of 33 polygenic scores, with a median percentage increase of 
`r median_pct_change`% in the respondents' generation (Appendix Table
\@ref(tab:tbl-inequality-effects)).  If inequalities in polygenic scores are
important for understanding social structure and mobility [@belsky2018genetic;
@Rimfeld_2018; @harden2021genetic], then these increases are substantive.
Similarly, since many polygenic scores are predictive of disease risk, they
could potentially increase health inequalities. In general, the evolutionary
history of anatomically modern humans is related to disease risk
[@benton2021influence]; understanding the role of contemporary natural selection
may help researchers to map the genetic architecture of health disparities.

```{r plot-mean-EA3-by-income, fig.align = "center", fig.cap = "Mean polygenic score for educational attainment (EA3) of children by household income group. Green is actual. Grey is hypothetical in the absence of selection effects.", fig.width = 4}
income_EA3 <- famhist %>% 
      filter(! is.na(n_children), ! is.na(income_cat)) %>% 
      mutate(
        `Household income` = factor(income_cat, 
                          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", 
                                      "> £100K"))
      ) %>% 
      group_by(`Household income`) %>% 
      summarize(
        `Without selection` = mean(EA3_excl_23andMe_UK, na.rm = TRUE),
        Actual              = weighted.mean(EA3_excl_23andMe_UK, n_children, 
                                              na.rm = TRUE)
      ) %>% 
      tidyr::pivot_longer(-`Household income`) %>% 
      rename(
        `Children's mean EA3` = value
      )

income_EA3 %>% 
              ggplot(aes(`Household income`, `Children's mean EA3`, fill = name)) +
                geom_col(position = "dodge") +
                scale_fill_manual(values = c("darkgreen", "grey65")) +
                theme(
                  legend.background = element_rect(
                                        fill = "white", 
                                        colour = "black",
                                        size = rel(0.5)
                                      ),
                  legend.margin = margin(t = 0, b = 5.5, l = 5.5, 
                                           r = 5.5),
                  legend.title = element_blank(), 
                  legend.position = c(.25, .85)
                )
```

@@

We run a standard mediation analysis in the framework of @baron1986moderator.
For each polygenic score where the bivariate correlation with RLRS is 
significant at $p$ < 0.05/33, we estimate 
\begin{align}
RLRS_i & = \alpha + \beta PGS_i + \gamma EA_i + X_i\mu + \varepsilon_i \\
EA_i & = \delta + \zeta PGS_i + X_i\mu + \eta_i
\end{align}
where $RLRS_i$ is relative lifetime reproductive success, $PGS_i$ is the
polygenic score, $EA_i$ is educational attainment (age of leaving fulltime
education), and $X_i$ is a vector of controls. The total effect of $PGS$ on
number of children is $\beta + \gamma \zeta$. The "indirect effect" mediated by
$EA$ is $\gamma \zeta$. The standard error of the indirect effect can be
calculated as

@@

where $\hat\sigma_\zeta$ is the standard error of $\hat\zeta$, etc.

We are not concerned about endogeneity of polygenic scores, since our theory
requires only that these correlate with human capital. However, endogeneity of
educational attainment is a concern, i.e., educational attainment might
correlate with unobserved characteristics that affect one's fertility (and which
are not related to human capital). To partially address this, we include
controls for age, sex, fluid IQ, height and risk attitude in $X$. Note that
controlling for fluid IQ is conservative, in the sense that IQ might also measure
human capital in the labour market; nevertheless, we control for it because 
fluid IQ might affect fertility decisions via other channels than human capital.

>>>>

```{r plot-res-wt, fig.cap = "Selection effects: weighted and unweighted regressions.  Each point represents a single bivariate regression of RLRS on a polygenic score. P value threshold is 0.05, Bonferroni-corrected for multiple comparisons. Confidence intervals are uncorrected."}

@@

res_mediation %<>% mutate(
                     # 0.025 not 0.05 because two-sided
                     sig     = abs(statistic_ind) > qnorm(1 - 0.025/
                                                            nrow(res_mediation)),
                     pos_sig = sig & prop_ind > 0,
                     bounded_ci = sign(estimate_total_conf_low) ==
                                  sign(estimate_total_conf_high)
                   )

@@

(ref:franzcite) [@franz2007ratios]

```{r plot-mediation, fig.align = "center", fig.cap = "Proportion of selection effect mediated by educational attainment, among polygenic scores with significant selection effects. Bootstrap confidence intervals for the proportion are shown only where the interval is bounded (ref:franzcite)."}

res_mediation <- res_mediation %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, prop_ind, order_abs(1)),
        `Indirect effect p < 0.05/23` = sig 
      )

res_mediation %>% 
      ggplot(aes(y = term)) + 
        geom_col(aes(x = prop_ind, fill = `Indirect effect p < 0.05/23`), 
                   width = .5) + 
        geom_errorbar(aes(
                          xmin = prop_ind_conf_low, 
                          xmax = prop_ind_conf_high,
                          linetype = "Bootstrap 95% c.i. uncorrected",
                        ),
                        data = res_mediation %>% filter(bounded_ci),
                        width = 0.2,
                        color = "grey20"
                      ) +
        scale_fill_manual(values = c("FALSE" = "grey65", "TRUE" = "steelblue4")) +
        my_vline +
        scale_x_continuous(labels = scales::percent, n.breaks = 7) +
        coord_cartesian(xlim = c(-0.5, 1)) +
        labs(x = "Proportion mediated", y = "") +
        scale_linetype_manual(name = "", 
          values = c("Bootstrap 95% c.i. uncorrected" = 1)) +
        my_theme(
          panel.grid.major.y = element_blank(),
          panel.grid.minor.x = element_blank(),
          legend.justification = "top"
        )

@@

Because selection effects are concentrated in lower-income groups, they may also
increase inequality with respect to polygenic scores. For example, Figure
\@ref(fig:plot-mean-EA3-by-income) plots mean polygenic scores for educational
attainment (EA3) among children from households of different income groups. The
blue bars show the actual means, i.e. parents' mean polygenic score weighted by
number of children. The grey bars show the hypothetical means if all households
had equal numbers of children. Natural selection against genes associated with
educational attainment is stronger at the bottom of the income distribution, and
this increases the differences between groups. Overall, natural selection
increases the correlation of polygenic scores with income for `r n_increases`
out of 33 polygenic scores, with a median percentage increase of 
`r median_pct_change`% in the respondents' generation (Appendix Table
\@ref(tab:tbl-inequality-effects)).  If inequalities in polygenic scores are
important for understanding social structure and mobility [@belsky2018genetic;
@Rimfeld_2018; @harden2021genetic], then these increases are substantive.
Similarly, since many polygenic scores are predictive of disease risk, they
could potentially increase health inequalities. In general, the evolutionary
history of anatomically modern humans is related to disease risk
[@benton2021influence]; understanding the role of contemporary natural selection
may help researchers to map the genetic architecture of health disparities.

```{r plot-mean-EA3-by-income, fig.align = "center", fig.cap = "Mean polygenic score for educational attainment (EA3) of children by household income group. Blue is actual. Grey is hypothetical in the absence of selection effects.", fig.width = 4}
income_EA3 <- famhist %>% 
      filter(! is.na(n_children), ! is.na(income_cat)) %>% 
      mutate(
        `Household income` = factor(income_cat, 
                          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", 
                                      "> £100K"))
      ) %>% 
      group_by(`Household income`) %>% 
      summarize(
        `Without selection` = mean(EA3_excl_23andMe_UK, na.rm = TRUE),
        Actual              = weighted.mean(EA3_excl_23andMe_UK, n_children, 
                                              na.rm = TRUE)
      ) %>% 
      tidyr::pivot_longer(-`Household income`) %>% 
      rename(
        `Children's mean EA3` = value
      )

income_EA3 %>% 
              ggplot(aes(`Household income`, `Children's mean EA3`, fill = name)) +
                geom_col(position = "dodge") +
                scale_fill_manual(values = c("steelblue4", "grey65")) +
                theme(
                  legend.background = element_rect(
                                        fill = "white", 
                                        colour = "black",
                                        size = rel(0.5)
                                      ),
                  legend.margin = margin(t = 0, b = 5.5, l = 5.5, 
                                           r = 5.5),
                  legend.title = element_blank(), 
                  legend.position = c(.25, .85)
                )
```

@@

We run a standard mediation analysis in the framework of @baron1986moderator.
For each polygenic score where the bivariate correlation with RLRS is 
significant at $p$ < 0.05/33, we estimate 
\begin{align}
RLRS_i & = \alpha + \beta PGS_i + \gamma EA_i + X_i\mu + \varepsilon_i \\
EA_i & = \delta + \zeta PGS_i + X_i\mu + \eta_i
\end{align}
where $RLRS_i$ is relative lifetime reproductive success, $PGS_i$ is the
polygenic score, $EA_i$ is educational attainment (age of leaving fulltime
education), and $X_i$ is a vector of controls. The total effect of $PGS$ on
$RLRS$ is $\beta + \gamma \zeta$. The "indirect effect" mediated by
$EA$ is $\gamma \zeta$. The standard error of the indirect effect can be
calculated as

@@

where $\hat\sigma_\zeta$ is the standard error of $\hat\zeta$, etc. We include 
controls for age and sex in $X$.
